---
title: "Project on ricci data"
author: "Dean Nasa 209381367"
date: "28 9 2022"
output:
  pdf_document: default
  html_document: default
---


In this project I analyzed the "RICCI" DATA - Data on firefighter promotion exams as part of the Ricci vs DeStafano court case to find if there is any racial discrimination between 3 types of firefighters (whites, Hispanics, blacks).

We would like to know if the expected weighted score (Combined) is different
between the position groups (Position) and the race (Race).
I Checked this using two-way analysis of variance, with integration 
(only if there is evidence in the data that it is needed).

*all tests were conducted with 0.05 confidence
```{r}
library(lawstat)
library(Stat2Data)
library(multcomp)
library(ggpubr)
library(rstatix)
library(tidyverse)
data(Ricci)
str(Ricci)



lm1 <- lm(Combine ~ Race*Position, Ricci)
# check to see if the data is ready 

# Normality of errors
plot(lm1)

print("hello world!")

# Heteroscedasticity

boxplot(Combine ~ Race*Position, Ricci)

# check for us - variance equality 

levene.test(Ricci$Combine, interaction(Ricci$Position ,Ricci$Race), location = "mean")

# check if the groups are balanced

table(Ricci$Position, Ricci$Race)

#we can see that the table is not balanced


#means of combination of groups

sum_of_groub_ab <- function(a,b){
  c <-0
  for(i in 1:length(Ricci$Race)){
    if((Ricci$Race[i] == a)&(Ricci$Position[i]== b)) {
      c<- c+1
    }
  }
  return(c)
}


mean_of_groub_ab <- function(a,b){
  s <-0
  for(i in 1:length(Ricci$Race)){
    if((Ricci$Race[i] == a)&(Ricci$Position[i]== b)) {
      s<- s +Ricci$Combine[i]
    }
  }
  return(s/sum_of_groub_ab(a,b))
}


mean(Ricci$Combine)
mean_of_groub_ab('W',"Captain")
mean_of_groub_ab('W',"Lieutenant")
mean_of_groub_ab('B',"Captain")
mean_of_groub_ab('B',"Lieutenant")
mean_of_groub_ab('H',"Captain")
mean_of_groub_ab('H',"Lieutenant")

aov_ric <-aov(Combine ~ Position * Race, data = Ricci)
summary(aov_ric)

# Only the race factor is significance.

# post hoc analysis

#bonforroni correction
pairwise <- glht(aov_ric, linfct = mcp(Race = 'Tukey'))
summary(pairwise, test = adjusted(type = "bonf"))

#tokey correction
tukey_hsd(aov_ric)


aov_ric2 <- aov(Combine ~ Race, data = Ricci)
summary(aov_ric2)

# We can see that the difference between Black and Hispanic is not significance.


```

In the above we focused on the sub-sample of the observations for which Lieutenant = Position holds and checked whether there is evidence that the assumption - Equality of differences does not hold for the weighted score (Combined) between the race groups. now I will perform a one-way analysis of variance and then also post hoc analyses.
```{r}
library(knitr)
library(lawstat)
library(Stat2Data)
library(dplyr)
data(Ricci)
str(Ricci)
Ricci_l <- Ricci[Ricci$Position == "Lieutenant",]
aov_l<-aov(Ricci_l$Combine~Ricci_l$Race)

# one-way analysis of variance examining whether in this sub-sample, Combined differs between the race groups


# pairwise comparison with bonferroni correction

pairwise.t.test(Ricci_l$Combine,Ricci_l$Race, p.adjust.method="bonferroni")

# we will reject h0 for white and black

# anova with tokey correction

TukeyHSD(aov(lm(Ricci_l$Combine ~ Ricci_l$Race)))
# we will reject h0 for white and black


# From the results in this part we concluded that there is discrimination against blacks in comparison to Whites

```


I can see that there is a difference between blacks, Hispanics and whites. and there is a difference between whites and blacks and there is a difference between whites and Hispanics. However there can be other explanations (for example, they come from different neighborhoods with different schools).

To conclude there is no doubt that there is a big problem with firefighter promotion exams. we cannot say precisely what is the source of the problem but the test discriminates hispanics and blacks in favor of the whites.